Staged Compilation With Two-Level Type Theory

### 2022-02-25 paper/paper.tex
stuff

In principle, it should be possible to have \emph{dependent memory layouts};
this would be a more advanced variation on type-passing polymorphism
\cite{intensional-poly}, where layouts of objects may depend on runtime data. For example,
length-prefixed flat arrays would not be a primitive notion, but simply defined
as $(n : \Nat) \times \msf{Vec}\,n\,A$. However, it seems to be highly
challenging to implement runtime systems and precise garbage collection in this
setting. See \texttt{sixten} \cite{sixten} for an experimental implementation of
dependent memory layouts.

@@

We proceed to summarize the key components of the model in a bit more detail. We
denote the components of the model by putting hats on 2LTT components, e.g.\ as
in $\hCon$. We summarize the key components in the following.

@@

The multi-stage calculus of Kawata and Igarashi \cite{multi-stage-calculus}
supports staging operations and type dependencies. However, there are major
differences to the current work. First, ibid.\ does not have universes or large
elimination, nor does it allow staged manipulation of type expressions. Second,
it does not support staging in the sense of Definition \ref{def:staging};
rather, it is a system for runtime code manipulation and evaluation. In other
words, programs may manipulate code fragments during execution, in more liberal
ways than in 2LTT, but there is no judgment which ensures that a purely
``object-level'' program can be extracted from a multi-stage program. We make a
disctinction between staged compilation and staged computation. For the former,
we expect staging in the sense of Definition \ref{def:staging}; this is
essential if we want to use staging as a lowering step in a compilation
pipeline. Staged computation on the other hand may support runtime code
construction and evaluation, but may not guarantee that these features can be
entirely computed away.

>>>>

%% In principle, it should be possible to have \emph{dependent memory layouts};
%% this would be a more advanced variation on type-passing polymorphism
%% \cite{intensional-poly}, where layouts of objects may depend on runtime data. For example,
%% length-prefixed flat arrays would not be a primitive notion, but simply defined
%% as $(n : \Nat) \times \msf{Vec}\,n\,A$. However, it seems to be highly
%% challenging to implement runtime systems and precise garbage collection in this
%% setting. See \texttt{sixten} \cite{sixten} for an experimental implementation of
%% dependent memory layouts.

@@

We proceed to summarize the key components of the model in a bit more detail. We
denote the components of the model by putting hats on 2LTT components, e.g.\ as
in $\hCon$.

@@

The multi-stage calculus of Kawata and Igarashi \cite{multi-stage-calculus}
supports staging operations and type dependencies. However, there are major
differences to the current work. First, ibid.\ does not have universes or large
elimination, nor does it allow staged manipulation of type expressions. Second,
it does not support staged compilation in the sense of Definition
\ref{def:staging}; rather, it is a system for runtime code manipulation and
evaluation (which we may call staged \emph{computation}). In other words,
programs may manipulate code fragments during execution, in more liberal ways
than in 2LTT, but there is no judgment which ensures that a purely
``object-level'' program can be extracted from a multi-stage program.

%% We make a
%% disctinction between staged compilation and staged computation. For the former,
%% we expect staging in the sense of Definition \ref{def:staging}; this is
%% essential if we want to use staging as a lowering step in a compilation
%% pipeline. Staged computation on the other hand may support runtime code
%% construction and evaluation, but may not guarantee that these features can be
%% entirely computed away.
